---
title: "Initial Modeling"
author: "Amber Potter & Nicholas Esposito"
date: "7/23/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries and data
```{r}
library(baseballr)
library(glue)
library(tidyverse)
library(knitr)
library(ranger)
```

```{r}
statcast_data <- read_rds("data/statcast_data_clean.rds")

names(statcast_data)[1] <- 'pitch_id'
```

# Filter out 2020 and 2022 seasons
 
```{r}
statcast_data_minus_2020_2022 <- statcast_data %>%
  filter(!(game_year %in% c(2020, 2022)))
```


# Reclassifying pitch type

```{r, warning = FALSE, message = FALSE}
statcast_data_minus_2020_2022 <- statcast_data_minus_2020_2022 %>%
  filter(!is.na(pitch_type)) %>%
  mutate(general_pitch_type = fct_recode(as.factor(pitch_type),
                                        "Breaking Ball" = "CU",
                                        "Fastball" = "FF",
                                        "Breaking Ball" = "SL",
                                        "Fastball" = "SI",
                                        "Changeup" = "CH",
                                        "Breaking Ball" = "KC",
                                        "Fastball" = "FC",
                                        "Fastball" = "FS",
                                        "Non-Pitch" = "IN",
                                        "Breaking Ball" = "FO",
                                        "Other" = "",
                                        "Changeup" = "EP",
                                        "Knuckle Ball" = "KN",
                                        "Non-Pitch" = "PO",
                                        "Fastball" = "FA",
                                        "Breaking Ball" = "SC",
                                        "Breaking Ball" = "CS"))
```

# Create an indicator for the last pitch of an at-bat

```{r}
statcast_data_minus_2020_2022 <- statcast_data_minus_2020_2022 %>%
  group_by(game_id, batter, at_bat_number) %>%
  #arrange(pitch_number) %>%
  mutate(pitch_count = max(pitch_number),
         last_pitch_of_at_bat = case_when(pitch_number == pitch_count ~ TRUE,
                                          TRUE ~ FALSE)) %>%
  ungroup()
```

# 200+ at bats (overall)

```{r}
statcast_data_minus_2020_2022 <- statcast_data_minus_2020_2022 %>%
  group_by(game_year, batter, stand) %>%
  mutate(total_player_at_bats = length(which(last_pitch_of_at_bat == TRUE))) %>%
  filter(total_player_at_bats >= 200)
```

## Strikeout labels

```{r}
statcast_data_minus_2020_2022 <- statcast_data_minus_2020_2022 %>%
  mutate(pitch_outcome = case_when(events != "" ~ events,
                                   events == "" ~ description)) %>%
  mutate(pitch_outcome = case_when(events == "strikeout" & description == "swinging_strike" ~ "strikeout_swinging",
                                   events == "strikeout" & description == "called_strike" ~ "strikeout_called",
                                   events == "strikeout" & !(description %in% c("swinging_strike", "called_strike")) ~ paste0("strikeout_other"),
                                   TRUE ~ pitch_outcome))
```


# Filter out NA values in each field alignment category, extreme outfield shift, and spring training/exhibition games

```{r}
statcast_data_minus_2020_2022 <- statcast_data_minus_2020_2022 %>%
  filter(if_fielding_alignment != "", !of_fielding_alignment %in% c("", "Extreme outfield shift"), 
         !game_type %in% c("S", "E")) # Filter NAs, ex. outfield shift, and ST/E games
```

# Filter out potential infield-in situations

```{r}
statcast_data_minus_2020_2022 <- statcast_data_minus_2020_2022 %>%
  filter(!(on_3b != "" & outs_when_up < 2 & if_fielding_alignment == "Shift"))
```

# Reduce shifting categories to shift vs standard

```{r}
statcast_data_minus_2020_2022 <- statcast_data_minus_2020_2022 %>%
  mutate(general_fielding_alignment = 
           case_when(if_fielding_alignment == "Infield shift" |
                       of_fielding_alignment == "4th outfielder" ~ "Shift",
                     TRUE ~ "Standard"))

# Reduce shifting subcategories to Standard vs Shift
statcast_data_minus_2020_2022 <- statcast_data_minus_2020_2022 %>%
  mutate(if_fielding_alignment = 
           case_when(if_fielding_alignment == "Infield shift" ~ "Shift",
                     if_fielding_alignment == "Strategic" |
                       if_fielding_alignment == "Standard"~ "Standard")) %>%
  mutate(of_fielding_alignment = 
           case_when(of_fielding_alignment == "4th outfielder" ~ "Shift",
                     of_fielding_alignment == "Strategic" |
                       of_fielding_alignment == "Standard" ~ "Standard"))
```

# Adjust BABIP calculation to exclude NAs and variables not involved in finding BABIP

```{r}
statcast_data_minus_2020_2022 <- statcast_data_minus_2020_2022 %>%
  mutate(babip_value_with_NAs = case_when(last_pitch_of_at_bat == FALSE | events == "sac_bunt_double_play" | events == "sac_bunt" | events == "home_run" | description != "hit_into_play" ~ NA_real_,
                                          TRUE ~ babip_value))
```


```{r}
statcast_data_minus_2020_2022 <- statcast_data_minus_2020_2022 %>%
#  filter(last_pitch_of_at_bat == TRUE) %>%
  mutate(walk = case_when(events == "walk" ~ 1,
                          TRUE ~ 0),
         strikeout = case_when(events == "strikeout" ~ 1,
                               TRUE ~ 0)) %>%
  group_by(game_year, batter, stand, if_fielding_alignment) %>%
  mutate(bb_rate = sum(walk) / length(which(last_pitch_of_at_bat == TRUE)),
            k_rate = sum(strikeout) / length(which(last_pitch_of_at_bat == TRUE)),
            babip = mean(babip_value_with_NAs, na.rm = TRUE),
            woba = sum(woba_value, na.rm = TRUE) / length(which(last_pitch_of_at_bat == TRUE)))
```

# Training set

```{r}
statcast_data_minus_2020_2022 <- statcast_data_minus_2020_2022 %>%
  mutate(batter_instance_identifier = paste(batter, game_year, stand, if_fielding_alignment, sep = ":")) 
    
statcast_training_data_rf_identifiers <- statcast_data_minus_2020_2022 %>%
  select(game_year, batter, stand, if_fielding_alignment, batter_instance_identifier) %>%
  unique() %>%
  group_by(game_year, stand, if_fielding_alignment) %>%
  sample_frac(.2)

statcast_training_data <- statcast_data_minus_2020_2022 %>%
  filter(batter_instance_identifier %in% statcast_training_data_rf_identifiers$batter_instance_identifier)
```


# At-bat level data

```{r}
statcast_data_at_bats_minus_2020_2022 <- statcast_data_minus_2020_2022 %>%
  filter(last_pitch_of_at_bat == TRUE)

unique(statcast_data_at_bats_minus_2020_2022$pitch_type)
```


## Variables as factors - SHOULD WE FORMAT MIDDLE NUMBERS (2,5,8) IN SAME WAY (M#?)

```{r}
statcast_data_at_bats_minus_2020_2022 <- statcast_data_at_bats_minus_2020_2022 %>%
  mutate(game_year = as.factor(game_year),
         batter = as.factor(batter),
         pitcher = as.factor(pitcher))
         #zone = as.factor(zone)) %>%
  mutate(stand = factor(stand, levels = c("R", "L")),
         p_throws = factor(p_throws, levels = c("R", "L")),
         if_fielding_alignment = factor(if_fielding_alignment, levels = c("Standard", "Shift")))
```

## Remove non-pitches

```{r}
statcast_data_at_bats_minus_2020_2022 <-statcast_data_at_bats_minus_2020_2022 %>%
  filter(general_pitch_type != "Non-Pitch")
```



## Fix zones

```{r}
statcast_data_at_bats_minus_2020_2022 <- statcast_data_at_bats_minus_2020_2022 %>%
  #mutate(zone = as.numeric(zone)) %>%
  mutate(zone_new = case_when(stand == "L" & zone %in% c(1, 4, 7) ~ zone + 2,
                          stand == "L" & zone %in% c(11, 13) ~ zone + 2,
                          stand == "L" & zone %in% c(3, 6, 9) ~ zone - 2,
                          stand == "L" & zone %in% c(12, 14) ~ zone - 2,
                          TRUE ~ zone)) %>%
  mutate(zone_simp = case_when(zone_new %in% c(1, 4, 7) ~ "Inside",
                               zone_new %in% c(2, 5, 8) ~ "Middle",
                               zone_new %in% c(3, 6, 9) ~ "Outside",
                               TRUE ~ "Out of Zone")) %>%
  select(- zone_new) %>%
  mutate(zone = as.factor(zone))

```

## Delta wOBA

```{r}
delta_woba_vector <- statcast_data_at_bats_minus_2020_2022 %>%
  group_by(game_year, batter, stand) %>%
  select(batter_instance_identifier, woba, total_player_at_bats) %>%
  unique() %>%
  mutate(count = n()) %>%
  filter(count == 2) %>%
  arrange(batter_instance_identifier) %>%
  # delta woba
  mutate(delta_woba = lead(woba) - woba) %>%
  mutate(delta_woba = ifelse(is.na(delta_woba), lag(delta_woba), delta_woba)) %>%
  mutate(batter_identifier = paste(batter, game_year, stand, sep = ":")) %>%
  # harmonic mean delta woba
  mutate(harmonic_mean_delta_woba = ((2 * total_player_at_bats * lead(total_player_at_bats)) / total_player_at_bats + lead(total_player_at_bats)) * delta_woba) %>%
  mutate(harmonic_mean_delta_woba = ifelse(is.na(harmonic_mean_delta_woba), lag(harmonic_mean_delta_woba), harmonic_mean_delta_woba)) %>%
  select(batter_identifier, delta_woba, harmonic_mean_delta_woba)

  

statcast_data_at_bats_minus_2020_2022 <- statcast_data_at_bats_minus_2020_2022 %>%
  mutate(batter_identifier = paste(batter, game_year, stand, sep = ":")) %>%
  left_join(delta_woba_vector, by = c("batter_identifier" = "batter_identifier", "game_year" = "game_year", "batter" = "batter", "stand" = "stand"))
```

## Binary At-Bat Outcome

```{r}
statcast_data_at_bats_minus_2020_2022 <- statcast_data_at_bats_minus_2020_2022 %>%
  mutate(on_base_binary = case_when(woba_value == 0 ~ 0,
                                    TRUE ~ 1))
```


## Write modified dataset as RDS

```{r}
write_rds(statcast_data_at_bats_minus_2020_2022, "data/statcast_at_bat_model_data.rds", compress = "gz")
```


# Basic Linear Model

```{r}
init_linear_model <- lm(woba_value ~ general_pitch_type + 
                          game_year + 
                          if_fielding_alignment + 
                          zone + 
                          stand+ 
                          p_throws+ 
                          woba + 
                          p_throws * stand * if_fielding_alignment + 
                          zone * if_fielding_alignment * general_pitch_type, 
                        data = statcast_data_at_bats_minus_2020_2022)


summary(init_linear_model)
```


When shifting takes place, how is wOBA affected?

control for shifting, batter info, pitcher info, handedness

Controlling for ability of hitter (batter's wOBA) -> this is the outcome on wOBA

mixed effects -> batter/pitcher

coeff. estimate for shifts

formal inference (use whole data for learning) 

pitch type driving changes? easier for righties to hit righty CBs

Use last pitch of at-bat (modeling at at-bat level!)