---
title: "joint-intermediate-analysis"
author: "Amber Potter"
date: "7/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load `baseballr` & Other Packages

```{r}
library(baseballr)
library(glue)
library(tidyverse)
library(knitr)
```

## Load Data

```{r}
statcast_data <- read_rds("data/statcast_data_clean.rds")
```

## Exploring Handedness Matchup Percentages

```{r}
statcast_data %>%
  count(stand, p_throws) %>%
  mutate(prop = n/sum(n))
```

## Percent missingness of shift variable by game type

```{r}
statcast_data %>%
  group_by(game_type) %>%
  count(na_if = if_fielding_alignment == "") %>%
  summarize(na_if, 
            prop_na_if_shift = n / sum(n)) %>%
  filter(is.na(na_if)) %>%
  arrange(prop_na_if_shift) %>%
  select(-na_if) %>%
  mutate(game_type = case_when(game_type == "W" ~ "World Series",
                               game_type == "L" ~ "League Championship",
                               game_type == "F" ~ "Wild Card",
                               game_type == "D" ~ "Divisional Series",
                               game_type == "R" ~ "Regular Season",
                               game_type == "S" ~ "Spring Training",
                               game_type == "E" ~ "Exhibition")) %>%
  kable(digits = 4, col.names = c("Game Type", "Percent Missing: IF Alignment"))
```

## Filter out Exhibition and ST games, and remove NA values from fielding alignments

- Based on percentage of missing values for each game type, we elected to exclude ST and Exhibition games. 

- Based on percentage of missing values for the regular season, all observations with NA for IF or OF fielding alignments are removed.

```{r}
statcast_data <- statcast_data %>%
  filter(if_fielding_alignment != "", of_fielding_alignment != "", !game_type %in% c("S", "E"))
```

## Shift vs Standard Alignment

### Runner on 3rd with Less than 2 Outs

- Proportion of plays where this scenario occurs

```{r}
statcast_data %>%
  mutate(runner_on_third = case_when(!is.na(on_3b) & (outs_when_up < 2) ~ TRUE,
                                      TRUE ~ FALSE)) %>%
  group_by(if_fielding_alignment, of_fielding_alignment) %>%
  summarize(prop = mean(runner_on_third)) 
```

- 
(following code is DUD)
```{r}
# not runner on third with less than 2 outs
# statcast_data %>%
#   mutate(runner_on_third = case_when(on_3b != "" & outs_when_up < 2 ~ TRUE,
#                                       TRUE ~ FALSE)) %>%
#   filter(runner_on_third == FALSE) %>%
#   group_by(batter) %>%
#   filter(!events == "") %>%
#   mutate(walk = case_when(events == "walk" ~ 1,
#                           TRUE ~ 0),
#          strikeout = case_when(events == "strikeout" ~ 1,
#                                TRUE ~ 0)) %>%
#   summarize(stand, p_throws, if_fielding_alignment, of_fielding_alignment, runner_on_third,
#           bb_rate = mean(walk),
#           k_rate = mean(strikeout),
#           babip = mean(babip_value),
#           woba = mean(woba_value)) %>%
#   select(-batter) %>%
#   group_by(stand, p_throws, if_fielding_alignment, of_fielding_alignment) %>%
#   summarize(bb_rate = mean(bb_rate), 
#             k_rate = mean(k_rate), 
#             babip = mean(babip), 
#             woba = mean(woba)) %>%
#   ungroup() %>%
#   unique()
# 
# # runner on third with less that 2 outs
# statcast_data %>%
#   mutate(runner_on_third = case_when(on_3b != "" & outs_when_up < 2 ~ TRUE,
#                                       TRUE ~ FALSE)) %>%
#   filter(runner_on_third == TRUE) %>%
#   group_by(batter) %>%
#   filter(!events == "") %>%
#   mutate(walk = case_when(events == "walk" ~ 1,
#                           TRUE ~ 0),
#          strikeout = case_when(events == "strikeout" ~ 1,
#                                TRUE ~ 0)) %>%
#   summarize(stand, p_throws, if_fielding_alignment, of_fielding_alignment, runner_on_third,
#           bb_rate = mean(walk),
#           k_rate = mean(strikeout),
#           babip = mean(babip_value),
#           woba = mean(woba_value)) %>%
#   select(-batter) %>%
#   group_by(stand, p_throws, if_fielding_alignment, of_fielding_alignment) %>%
#   summarize(bb_rate = mean(bb_rate), 
#             k_rate = mean(k_rate), 
#             babip = mean(babip), 
#             woba = mean(woba)) %>%
#   ungroup() %>%
#   unique()
```

## Threshold for plate appearances

- We plotted the counts for number of at-bats

```{r}
statcast_data %>%
  group_by(batter) %>%
  mutate(total_player_at_bats = length(which(events != ""))) %>%
  ggplot() +
  geom_histogram(aes(x = total_player_at_bats))
```


## Statistics for infield-in scenarios

-Strikeout rate

-Walk rate

-BABIP

-WOBA

```{r}
statcast_data %>%
  # situation identifier
  mutate(runner_on_third = case_when(on_3b != "" & outs_when_up < 2 ~ TRUE,
                                      TRUE ~ FALSE)) %>%
  # filtering for 400+ at bats
  group_by(batter) %>%
  mutate(total_player_at_bats = length(which(events != ""))) %>%
  filter(total_player_at_bats >= 400) %>%
  
  # effectively counting each at bat once
  filter(!events == "") %>%
  # was the outcome of at bat a walk or strikeout
  mutate(walk = case_when(events == "walk" ~ 1,
                          TRUE ~ 0),
         strikeout = case_when(events == "strikeout" ~ 1,
                               TRUE ~ 0)) %>%
  # summarizing statistics by batter in and out of this situation
  group_by(batter, runner_on_third) %>%
  summarize(runner_on_third,
            bb_rate = mean(walk),
        k_rate = mean(strikeout),
        babip = mean(babip_value),
        woba = mean(woba_value),
        .groups = "drop") %>%
  # only including each batter once in and out of this situation
  unique() %>%
  # summarizing offensive stats based on runner on third scenarios
  group_by(runner_on_third) %>%
  summarize(bb_rate = mean(bb_rate), 
          k_rate = mean(k_rate), 
          babip = mean(babip), 
          woba = mean(woba))
```


- Confidence Intervals for infield-in scenarios

```{r}
runner_on_third_data <- statcast_data %>%
  # situation identifier
  mutate(runner_on_third = case_when(on_3b != "" & outs_when_up < 2 ~ TRUE,
                                      TRUE ~ FALSE)) %>%
  # filtering for 400+ at bats
  group_by(batter) %>%
  mutate(total_player_at_bats = length(which(events != ""))) %>%
  filter(total_player_at_bats >= 400) %>%
    # effectively counting each at bat once
  filter(!events == "") %>%
  # was the outcome of at bat a walk or strikeout
  mutate(walk = case_when(events == "walk" ~ 1,
                          TRUE ~ 0),
         strikeout = case_when(events == "strikeout" ~ 1,
                               TRUE ~ 0)) %>%
  # summarizing statistics by batter in and out of this situation
  group_by(batter, runner_on_third) %>%
  summarize(runner_on_third,
            bb_rate = mean(walk),
        k_rate = mean(strikeout),
        babip = mean(babip_value),
        woba = mean(woba_value),
        .groups = "drop") %>%
  # only including each batter once in and out of this situation
  unique() %>%
  filter(runner_on_third == TRUE)

no_runner_on_third_data <- statcast_data %>%
  # situation identifier
  mutate(runner_on_third = case_when(on_3b != "" & outs_when_up < 2 ~ TRUE,
                                      TRUE ~ FALSE)) %>%
  # filtering for 400+ at bats
  group_by(batter) %>%
  mutate(total_player_at_bats = length(which(events != ""))) %>%
  filter(total_player_at_bats >= 400) %>%
    # effectively counting each at bat once
  filter(!events == "") %>%
  # was the outcome of at bat a walk or strikeout
  mutate(walk = case_when(events == "walk" ~ 1,
                          TRUE ~ 0),
         strikeout = case_when(events == "strikeout" ~ 1,
                               TRUE ~ 0)) %>%
  # summarizing statistics by batter in and out of this situation
  group_by(batter, runner_on_third) %>%
  summarize(runner_on_third,
            bb_rate = mean(walk),
        k_rate = mean(strikeout),
        babip = mean(babip_value),
        woba = mean(woba_value),
        .groups = "drop") %>%
  # only including each batter once in and out of this situation
  unique() %>%
  filter(runner_on_third == FALSE)

runner_on_third_conf_int <- lm(woba ~ 1, runner_on_third_data)
confint(runner_on_third_conf_int, level = 0.95)

no_runner_on_third_conf_int <- lm(woba ~ 1, no_runner_on_third_data)
confint(no_runner_on_third_conf_int, level = 0.95)
```


## Percentages of standard and shifted IF and OF arangements based on batter and pitcher handedness

- Grouping into general shifted and standard categories

```{r}
statcast_data <- statcast_data %>%
  mutate(general_fielding_alignment = 
           case_when(if_fielding_alignment == "Infield shift" |
                       of_fielding_alignment == "4th outfielder" |
                       of_fielding_alignment == "Extreme outfield shift" ~
                       "Shift",
                     TRUE ~ "Standard"))


statcast_data <- statcast_data %>%
  mutate(if_fielding_alignment = 
           case_when(if_fielding_alignment == "Infield shift" ~ "Shift",
                     if_fielding_alignment == "Strategic" |
                       if_fielding_alignment == "Standard"~ "Standard")) %>%
  mutate(of_fielding_alignment = 
           case_when(of_fielding_alignment == "4th outfielder" |
                       of_fielding_alignment == "Extreme outfield shift"~ "Shift",
                     of_fielding_alignment == "Strategic" |
                       of_fielding_alignment == "Standard" ~ "Standard"))
```

- Viewing percentages of general fielding arangements based on handedness

```{r}
statcast_data %>%
  count(stand, p_throws, if_fielding_alignment, of_fielding_alignment) %>%
  mutate(prop = n/sum(n))
```

## Creating Summary Statistics

- Strikeout rate

- Walk rate

- BABIP

- WOBA

## Average Summary Stats within handedness and fielding alignment groups

```{r}
statcast_data %>%
  group_by(batter) %>%
  mutate(total_player_at_bats = length(which(events != ""))) %>%
  filter(total_player_at_bats >= 400) %>%
  filter(!events == "") %>%
  mutate(walk = case_when(events == "walk" ~ 1,
                          TRUE ~ 0),
         strikeout = case_when(events == "strikeout" ~ 1,
                               TRUE ~ 0)) %>%
  group_by(batter, general_fielding_alignment) %>%
  summarize(stand, p_throws,
            bb_rate = mean(walk),
            k_rate = mean(strikeout),
            babip = mean(babip_value),
            woba = mean(woba_value),
            .groups = "drop") %>% 
  unique() %>%
  select(-batter) %>%
  group_by(stand, general_fielding_alignment) %>%
  summarize(bb_rate = mean(bb_rate), 
            k_rate = mean(k_rate), 
            babip = mean(babip), 
            woba = mean(woba),
            .groups = "drop")
```



## Distribution of WOBA for different fielding alignment


```{r}
statcast_data %>%
  group_by(batter) %>%
  mutate(total_player_at_bats = length(which(events != ""))) %>%
  filter(total_player_at_bats >= 400) %>%
  filter(!events == "") %>%
  group_by(batter, general_fielding_alignment) %>%
  summarize(stand, p_throws,
            woba = mean(woba_value),
            .groups = "drop") %>% 
  unique() %>%
  group_by(general_fielding_alignment) %>%
  summarize(woba = mean(woba),
            .groups = "drop")
  ggplot(aes(x = woba)) +
  geom_histogram() +
  facet_wrap( ~ general_fielding_alignment,)
```



woba is based on all players at bats, not just when player is shifted