---
title: "Determining the Critical Influences Affecting Offensive Outcomes of Shifted At-Bats"
author: "Nicholas Esposito and Amber Potter"
date: "7/22/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

Through our work, we hope to:

- Gain an understanding of the situations in which shifting occurs

- Analyze the differences within each subcategory of shifting 

- Determine the influential factors involved in deciding to shift and the offensive outcomes of shifted plays.

We also have two main questions that will guide our work:

- What factors contribute to the changes in offensive outcomes, based on batter handedness, in shifted situations? 

- Based on these factors, when is it most advantageous to shift?

By looking into this, our overarching goal is to provide insight to aid in-game and season-long decisions and to better understand the impact of shifting on the game of baseball.

## Data

For our work, we accessed pitch-by-pitch statcast data from the baseballR package (Bill Petti). This dataset contained pitch-level information dating back to 2015, when statcast technology was introduced in all 30 MLB stadiums, up to July 13th, 2022, which we declared the current season’s cutoff. The initial dataset had 93 variables per observation; however, during preprocessing, we performed several transformations to clean the dataset:

- Removed all columns with only NA values (empty columns)

- Grouped event and description columns for different types of strikeouts, so strikeout descriptions in the resulting column include whether the strikeout was swinging or looking.

- Added an at-bat column to keep track of the number of at-bats per player per game (at-bat #1, #2, etc.).

After preprocessing, each observation had 88 variables, such as:
- Game Year, Game Date, Game Type (Spring Training, Regular Season)

- Pitcher and Batter Identifiers (Name, Unique ID, Handiness)

- Game Information (Score, Outs, Runners On-Base, Inning)

- Pitch Information (Pitch Type, Pitch Location, Pitch Outcome)

- Positional (Shifting) Data

- Player Statistics (wOBA, BABIP)

During initial EDA, we discovered a high proportion of missing shift values in the Spring Training pitch data (~78%), as well as a low proportion of missing shift values in the Regular Season (~0.01%). From this, we chose to exclude the Spring Training data from each year, and all observations with missing shift values in either the infield or outfield alignment column. 

In situations with a runner on third base and less than two outs, the infield is likely to be shifted in. The strategic shift category may contain a significant proportion of these scenarios - about 17% of strategic-standard and strategic-strategic observations (infield-outfield). Such situations often inflate offensive outcomes, as the infield-in shift promotes variance in run scoring to maximize win probability at the expense of expected runs allowed. To adjust for this, we removed observations in which the infield would be shifted in. However, to ensure we were not excluding situations in which there was a runner on third base with less than two outs but the infield was not shifted in, we only removed the observations marked as “shift”. 

## EDA

### Load `baseballr` & Other Packages

```{r}
library(baseballr)
library(glue)
library(tidyverse)
library(knitr)
```

### Load Data

```{r}
statcast_data <- read_rds("data/statcast_data_clean.rds")

names(statcast_data)[1] <- 'pitch_id'
```

### End of at bat column

```{r}
statcast_data <- statcast_data %>%
  group_by(game_id, batter, at_bat_number) %>%
  #arrange(pitch_number) %>%
  mutate(pitch_count = max(pitch_number),
         last_pitch_of_at_bat = case_when(pitch_number == pitch_count ~ TRUE,
                                          TRUE ~ FALSE)) %>%
  ungroup()
```

### Exploring Handedness Matchup Percentages

```{r}
statcast_data %>%
  count(stand, p_throws) %>%
  mutate(prop = n/sum(n))
```

### Percent missingness of shift variable by game type

```{r}
statcast_data %>%
  group_by(game_type) %>%
  count(na_if = if_fielding_alignment == "") %>%
  summarize(na_if, 
            prop_na_if_shift = n / sum(n)) %>%
  filter(is.na(na_if)) %>%
  arrange(prop_na_if_shift) %>%
  select(-na_if) %>%
  mutate(game_type = case_when(game_type == "W" ~ "World Series",
                               game_type == "L" ~ "League Championship",
                               game_type == "F" ~ "Wild Card",
                               game_type == "D" ~ "Divisional Series",
                               game_type == "R" ~ "Regular Season",
                               game_type == "S" ~ "Spring Training",
                               game_type == "E" ~ "Exhibition")) %>%
  kable(digits = 4, col.names = c("Game Type", "Percent Missing: IF Alignment"))
```

### Filter out Exhibition games, ST games, Extreme Outfield Shift, and NA values from fielding alignments

- Based on percentage of missing values for each game type, we elected to exclude ST and Exhibition games. 

- Based on percentage of missing values for the regular season, all observations with NA for IF or OF fielding alignments are removed.

- Based on limited values for the extreme outfield shift, all observations with this alignment for IF or OF fielding alignments are removed.

```{r}
statcast_data <- statcast_data %>%
  filter(if_fielding_alignment != "", !of_fielding_alignment %in% c("", "Extreme outfield shift"), 
         !game_type %in% c("S", "E")) # Filter NAs, ex. outfield shift, and ST/E games
```

## Shift vs Standard Alignment - Runner on 3rd with Less than 2 Outs

- Proportion of plays where this scenario occurs

```{r}
statcast_data %>%
  mutate(runner_on_third = case_when(!is.na(on_3b) & (outs_when_up < 2) ~ TRUE,
                                      TRUE ~ FALSE)) %>%
  filter(last_pitch_of_at_bat == TRUE) %>%
  group_by(if_fielding_alignment, of_fielding_alignment) %>%
  summarize(prop = mean(runner_on_third)) %>% # Prop. for runners on third in each shifting scenario
  kable()
```

## Statistics for infield-in scenarios

- Strikeout rate

- Walk rate

- BABIP

- WOBA

```{r}
statcast_data %>%
  # situation identifier
  mutate(runner_on_third = case_when(on_3b != "" & outs_when_up < 2 ~ TRUE,
                                      TRUE ~ FALSE)) %>%
  mutate(babip_value_with_NAs = case_when(last_pitch_of_at_bat == FALSE | events == "sac_bunt_double_play" | events == "sac_bunt" | events == "home_run" | description != "hit_into_play" ~ NA_real_,
                                          TRUE ~ babip_value)) %>%
  # filtering for 400+ at bats
  group_by(game_year, batter) %>%
  mutate(total_player_at_bats = length(which(last_pitch_of_at_bat == TRUE))) %>%
  filter(total_player_at_bats >= 200) %>%
  
  # effectively counting each at bat once
  filter(last_pitch_of_at_bat == TRUE) %>%
  # was the outcome of at bat a walk or strikeout
  mutate(walk = case_when(events == "walk" ~ 1,
                          TRUE ~ 0),
         strikeout = case_when(events == "strikeout" ~ 1,
                               TRUE ~ 0)) %>%
  # summarizing statistics by batter in and out of this situation
  group_by(game_year, batter, runner_on_third) %>%
  summarize(runner_on_third,
            bb_rate = mean(walk),
        k_rate = mean(strikeout),
        babip = mean(babip_value_with_NAs, na.rm = TRUE),
        woba = mean(woba_value, na.rm = TRUE),
        .groups = "drop") %>%
  # only including each batter once in and out of this situation
  unique() %>%
  # summarizing offensive stats based on runner on third scenarios
  group_by(game_year, runner_on_third) %>%
  summarize(bb_rate = mean(bb_rate), 
          k_rate = mean(k_rate), 
          babip = mean(babip), 
          woba = mean(woba)) %>%
  kable()
```

- Confidence Intervals for Infield-in Scenarios

```{r}
runner_on_third_data <- statcast_data %>%
  # situation identifier
  mutate(runner_on_third = case_when(on_3b != "" & outs_when_up < 2 ~ TRUE,
                                      TRUE ~ FALSE)) %>%
    mutate(babip_value_with_NAs = case_when(last_pitch_of_at_bat == FALSE | events == "sac_bunt_double_play" | events == "sac_bunt" | events == "home_run" | description != "hit_into_play" ~ NA_real_,
                                          TRUE ~ babip_value)) %>%
  # filtering for 400+ at bats
  group_by(game_year, batter) %>%
  mutate(total_player_at_bats = length(which(last_pitch_of_at_bat == TRUE))) %>%
  filter(total_player_at_bats >= 400) %>%
    # effectively counting each at bat once
  filter(last_pitch_of_at_bat == TRUE) %>%
  # was the outcome of at bat a walk or strikeout
  mutate(walk = case_when(events == "walk" ~ 1,
                          TRUE ~ 0),
         strikeout = case_when(events == "strikeout" ~ 1,
                               TRUE ~ 0)) %>%
  # summarizing statistics by batter in and out of this situation
  group_by(game_year, batter, runner_on_third) %>%
  summarize(runner_on_third,
            bb_rate = mean(walk),
        k_rate = mean(strikeout),
        babip = mean(babip_value),
        woba = mean(woba_value),
        .groups = "drop") %>%
  # only including each batter once in and out of this situation
  unique() %>%
  filter(runner_on_third == TRUE)

no_runner_on_third_data <- statcast_data %>%
  # situation identifier
  mutate(runner_on_third = case_when(on_3b != "" & outs_when_up < 2 ~ TRUE,
                                      TRUE ~ FALSE)) %>%
    mutate(babip_value_with_NAs = case_when(last_pitch_of_at_bat == FALSE | events == "sac_bunt_double_play" | events == "sac_bunt" | events == "home_run" | description != "hit_into_play" ~ NA_real_,
                                          TRUE ~ babip_value)) %>%
  # filtering for 400+ at bats
  group_by(game_year, batter) %>%
  mutate(total_player_at_bats = length(which(last_pitch_of_at_bat == TRUE))) %>%
  filter(total_player_at_bats >= 400) %>%
    # effectively counting each at bat once
  filter(last_pitch_of_at_bat == TRUE) %>%
  # was the outcome of at bat a walk or strikeout
  mutate(walk = case_when(events == "walk" ~ 1,
                          TRUE ~ 0),
         strikeout = case_when(events == "strikeout" ~ 1,
                               TRUE ~ 0)) %>%
  # summarizing statistics by batter in and out of this situation
  group_by(game_year, batter, runner_on_third) %>%
  summarize(runner_on_third,
            bb_rate = mean(walk),
        k_rate = mean(strikeout),
        babip = mean(babip_value),
        woba = mean(woba_value),
        .groups = "drop") %>%
  # only including each batter once in and out of this situation
  unique() %>%
  filter(runner_on_third == FALSE)

runner_on_third_conf_int <- lm(woba ~ 1, runner_on_third_data)
confint(runner_on_third_conf_int, level = 0.95)

no_runner_on_third_conf_int <- lm(woba ~ 1, no_runner_on_third_data)
confint(no_runner_on_third_conf_int, level = 0.95) %>%
  kable()
```

## Filter out all Infield-in Scenarios (runner on third, less than two outs)

```{r}
statcast_data <- statcast_data %>%
  filter(!(on_3b != "" & outs_when_up < 2))
```

## Percentages of Standard and Shifted IF and OF Alignments Based on Batter Handedness

- Grouping into general shifted and standard categories

```{r}
statcast_data <- statcast_data %>%
  mutate(general_fielding_alignment = 
           case_when(if_fielding_alignment == "Infield shift" |
                       of_fielding_alignment == "4th outfielder" ~ "Shift",
                     TRUE ~ "Standard"))

# Reduce shifting subcategories to Standard vs Shift
statcast_data <- statcast_data %>%
  mutate(if_fielding_alignment = 
           case_when(if_fielding_alignment == "Infield shift" ~ "Shift",
                     if_fielding_alignment == "Strategic" |
                       if_fielding_alignment == "Standard"~ "Standard")) %>%
  mutate(of_fielding_alignment = 
           case_when(of_fielding_alignment == "4th outfielder" ~ "Shift",
                     of_fielding_alignment == "Strategic" |
                       of_fielding_alignment == "Standard" ~ "Standard"))
```

- Viewing percentages of general fielding arangements based on handedness

```{r}
statcast_data %>%
  count(stand, p_throws, if_fielding_alignment, of_fielding_alignment) %>%
  mutate(prop = n/sum(n)) 
```

## Creating Summary Statistics

- Strikeout rate

- Walk rate

- BABIP

- WOBA

## Average Summary Stats within Batter Handedness and Shift (Since 2015)

```{r}
statcast_data %>%
  group_by(game_year, batter) %>%
  mutate(total_player_at_bats = length(which(last_pitch_of_at_bat == TRUE))) %>%
  mutate(babip_value_with_NAs = case_when(last_pitch_of_at_bat == FALSE | events == "sac_bunt_double_play" | events == "sac_bunt" | events == "home_run" | description != "hit_into_play" ~ NA_real_,
                                          TRUE ~ babip_value)) %>%
  filter(total_player_at_bats >= 200) %>%
  filter(last_pitch_of_at_bat == TRUE) %>%
  mutate(walk = case_when(events == "walk" ~ 1,
                          TRUE ~ 0),
         strikeout = case_when(events == "strikeout" ~ 1,
                               TRUE ~ 0)) %>%
  group_by(game_year, batter, stand, if_fielding_alignment) %>%
  summarize(bb_rate = mean(walk),
            k_rate = mean(strikeout),
            babip = mean(babip_value_with_NAs, na.rm = TRUE),
            woba = mean(woba_value, na.rm = TRUE),
            .groups = "drop") %>% 
  unique() %>%
  select(-batter) %>%
  group_by(game_year, stand, if_fielding_alignment) %>%
  summarize(bb_rate = mean(bb_rate), 
            k_rate = mean(k_rate), 
            babip = mean(babip, na.rm = TRUE), 
            woba = mean(woba, na.rm = TRUE),
            .groups = "drop") %>%
  kable()
```

## Distribution of wOBA for Different Fielding Alignment

```{r}
statcast_data %>%
  group_by(game_year, batter) %>%
  mutate(total_player_at_bats = length(which(last_pitch_of_at_bat == TRUE))) %>%
  filter(total_player_at_bats >= 200) %>%
  filter(last_pitch_of_at_bat == TRUE) %>%
  group_by(game_year, batter, if_fielding_alignment) %>%
  summarize(stand, p_throws,
            woba = mean(woba_value, na.rm = TRUE),
            .groups = "drop") %>% 
  unique() %>%
  ggplot(aes(x = woba)) +
  geom_histogram() +
  facet_grid(game_year ~ if_fielding_alignment) +
  labs(x = "wOBA",
       y = "Frequency",
       title = "wOBA Distributions have Greater Variability when Against the Shift") +
  theme_minimal()
```

## Methodology

- Our primary tool for analysis is mixed effects modeling, as this approach will provide the best   insight on what factors contribute to the changes in offensive outcome. The wOBA statistic     (weighted on-base average) is very similar to on-base percentage, although it accounts for how a player got on base. Given that the goal of the shift is to keep runners off base, using a measure that encapsulates how players get on is a great indicator of shift success. Thus, wOBA will serve as our response variable for modeling. 

- To estimate the effects of the shift, we will control for the batter using overall wOBA totals (fixed effect), which will regress players towards the mean.

## Results

- TBD

## Discussion - Limitations and Future Work

- We encountered several limitations with our exploratory work:

  - We elected to exclude observations from both 2020 and 2022, as both years did not include a full     season's worth of pitch data, and thus did not sufficiently represent year-long trends. 
  
  - The shifting categories are very broad and the differences between shift groupings is not           distinct. Similarly, we did not have player positional information. Both of these limited our       overall understanding of the specifics of each shift. 

- For future work, we have considered within-batter matching (more specifically, propensity score     matching) to further enhance our understanding of shifting trends and tendencies. This work would   look into the probability of a shift taking place, thus better helping in-game decisions.                                                               
## Acknowledgements

- We would like to thank both Dr. Adam Brodie and Dr. Ron Yurko for their help and guidance throughout this process. 

## References

https://batflipsandnerds.com/2021/02/03/the-psychology-of-the-infield-shift/

https://billpetti.github.io/baseballr/reference/index.html

https://billpetti.github.io/baseballr/

https://www.mlb.com/glossary/statcast/shifts